---
title: "Bikeshare SpaceTime Prediction"
subtitle: "CPLN 5080 - Spring 2026"
author: "Veronica Zheng"
date: today
format:
  html:
    code-fold: true
    code-tools: true
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

# 1. Introduction

Philadelphia's Indego bike share system serves a growing number of riders across the city, and accurately predicting station-level demand is essential for effective rebalancing and fleet management. This report applies a space-time panel regression framework to forecast hourly trip counts at each station using Q1 and Q2 2025 trip data. The analysis builds on temporal lag features, weather conditions, demographic context, and engineered seasonal indicators to capture the complex patterns that drive bike share demand.

Q1 2025 covers January through March, a period defined by cold temperatures and low recreational ridership. Q2 2025 covers April through June, when warmer weather, university semester endings, and increased outdoor activity reshape demand. Together, the two quarters offer a meaningful comparison for assessing whether models trained on one season generalize to another.

The workflow follows five OLS regression models of increasing complexity, evaluated using Mean Absolute Error (MAE), Root Mean Squared Error (RMSE), and R-squared. A Poisson regression model is also included as an alternative suited to count data. The best-performing baseline model is then extended with three new engineered features specific to Q2 demand patterns. Error analysis examines how model accuracy varies by time of day, day of week, station, and neighborhood demographics. A final reflection considers the equity and operational implications of deploying these models at scale.

# 2. Report

## Study Area and Data

This analysis uses Indego bike share trip records for Q1 2025 (January through March) and Q2 2025 (April through June) in Philadelphia. Station-level census data was joined using tidycensus, drawing on tract-level estimates of median household income, transit commute share, and racial composition. Weather data was fetched hourly from the Iowa Environmental Mesonet API for Philadelphia International Airport (KPHL), capturing temperature, precipitation, and wind speed.

Trip counts were aggregated to the station-hour level, creating a panel dataset where each row represents one station in one hour. Temporal lag features (one hour, three hours, and one day prior) capture autocorrelation in demand. A temporal train-test split was used within each quarter: earlier weeks served as training data and later weeks as the test set.

## Model Performance

Five OLS regression models were evaluated for both quarters. Model complexity increased progressively, beginning with just time-of-day and day-of-week predictors and ending with station fixed effects, weather variables, temporal lags, and demographic controls. Model 4, which combined all predictors except demographic interactions, achieved the best balance of accuracy and generalization, with a Q1 test MAE of approximately 0.69 and a Q2 test MAE of approximately 1.07.

Q2 showed consistently higher errors than Q1 across all models. This reflects greater variability in spring and early summer demand driven by recreational trips, event-related surges, and semester transitions that are harder to capture with general features. The Poisson regression model, implemented using fastglm for computational efficiency, performed comparably to the best OLS model and is better suited to count data in theory.



## Feature Engineering

Three new features were added to Model 4 for the Q2 analysis. The first is a binary indicator for perfect cycling weather, defined as temperatures between 60 and 75 degrees Fahrenheit with no precipitation. The second captures the university semester end period from late April through mid-May, when student ridership is elevated and then drops sharply. The third flags the Memorial Day holiday period, which shows a distinct recreational demand spike distinct from a normal Monday. Together these features reduced Q2 test MAE from approximately 1.07 to approximately 0.94.

## Error Analysis

Prediction errors were largest during morning and evening commute hours, when demand spikes are steep and brief. Weekend afternoons also showed higher errors, likely due to recreational trips that are less regular than commutes. Geographically, stations near university campuses and Center City had the most underprediction during event-driven surges. Stations in lower-income neighborhoods with limited transit alternatives tended to have smaller absolute errors but also lower overall ridership, suggesting the model may underserve those areas if used for rebalancing decisions.


## Reflection

A transit agency deploying this model would need to account for its systematic underprediction at high-demand times and locations. Equity implications are significant because optimizing for global MAE may concentrate rebalancing effort on busy high-income corridors while leaving lower-ridership stations underserved. Future work could explore separate models by station cluster or time-of-day category, or incorporate real-time data feeds to reduce lag in predictions. The model as built provides a useful demand signal but should be paired with human judgment and equity-aware rebalancing policies.

# 3. QMD Link

[GitHub Link to QMD](https://github.com/vrnczay/CPLN-5920-Portfolio/blob/main/labs/lab_5/week11_inclass_and_lab5.qmd)


# 4. Technical Appendix

## 4.1 Q1 2025 In-Class Analysis

```{r setup}
#| message: false
#| warning: false
#| include: false

Sys.setlocale("LC_TIME", "English") 


options(scipen = 999)

library(tidyverse)
library(lubridate)
library(sf)
library(tigris)
library(tidycensus)
library(riem)  # For Philadelphia weather from ASOS stations

library(viridis) # color scales for visualization
library(knitr)
library(kableExtra)
library(patchwork)
library(glmmTMB)

library(here)
options(scipen = 999) # Get rid of scientific notation to make it look tidier

```

```{r themes}
#| include: false
plotTheme <- theme_minimal() +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption  = element_text(size = 8),
    axis.text.x   = element_text(size = 10, angle = 45, hjust = 1),
    axis.text.y   = element_text(size = 10),
    axis.title    = element_text(size = 11, face = "bold"),
    panel.grid.major = element_line(colour = "#D0D0D0", linewidth = 0.2),
    panel.grid.minor = element_blank(),
    axis.ticks    = element_blank(),
    legend.position = "right"
  )

mapTheme <- theme_void() +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption  = element_text(size = 8),
    legend.position  = "right",
    plot.margin      = margin(1, 1, 1, 1, "cm"),
    legend.key.height = unit(1, "cm"),
    legend.key.width  = unit(0.2, "cm")
  )

palette5 <- c("#eff3ff", "#bdd7e7", "#6baed6", "#3182bd", "#08519c")
```

```{r set-census-api-key}
#| include: false

census_api_key(Sys.getenv("CENSUS_API_KEY"))

#Use the following code if you're not able to access the key in your environment
#census_api_key("yourkeyhere", install=TRUE, overwrite=TRUE)
```

### Load and Process Indego Trip Data (Q1 2025)

```{r load_indego}
# Read Q1 2025 data
indego <- read_csv(here("data/indego-trips-2025-q1.csv"))
```

```{r explore_data}
# How many trips?
cat("Total trips in Q1 2025:", nrow(indego), "\n")

# Date range
cat("Date range:", 
    min(mdy_hm(indego$start_time)), "to", 
    max(mdy_hm(indego$start_time)), "\n")

# How many unique stations?
cat("Unique start stations:", length(unique(indego$start_station)), "\n")

# Trip types
table(indego$trip_route_category)

# Passholder types
table(indego$passholder_type)

# Bike types
table(indego$bike_type)
```

```{r create-time-bins}
indego <- indego %>%
  mutate(
    # Parse datetime
    start_datetime = mdy_hm(start_time),
    end_datetime = mdy_hm(end_time),
    
    # Create hourly bins
    interval60 = floor_date(start_datetime, unit = "hour"),
    
    # Extract time features
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    
    # Create useful indicators
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

# Look at temporal features
head(indego %>% select(start_datetime, interval60, week, dotw, hour, weekend))
```

`interval60` rounds each trip's start timestamp down to the nearest hour using `floor_date()`. For example, a trip that started at 8:47 AM gets assigned to the 8:00 AM bin, and a trip at 8:03 AM also gets assigned to 8:00 AM. This creates a consistent hourly time unit that lets us aggregate all trips happening within the same hour at the same station into a single count, which is the foundation of our space-time panel structure.

#### Daily Trip

```{r trips_over_time}
#| message: false
#| warning: false

# Daily trip counts
daily_trips <- indego %>%
  group_by(date) %>%
  summarize(trips = n())

ggplot(daily_trips, aes(x = date, y = trips)) +
  geom_line(color = "#3182bd", linewidth = 1) +
  geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
  labs(
    title = "Indego Daily Ridership - Q1 2025",
    subtitle = "Winter demand patterns in Philadelphia",
    x = "Date",
    y = "Daily Trips",
    caption = "Source: Indego bike share"
  ) +
  plotTheme
```

Ridership shows a clear upward trend from January through March, rising from roughly 1,500 trips per day in early January to over 3,500 by late March, reflecting the seasonal effect of warming temperatures making cycling more appealing. There is also a strong weekly cyclical pattern throughout the quarter, with regular peaks and troughs corresponding to weekdays versus weekends. The most notable anomaly is the sharp spike on February 14th, which far exceeds any typical Friday ridership — this was the day of the Philadelphia Eagles Super Bowl victory parade, which drew massive crowds downtown and drove an unusual surge in bike share usage.

#### Hourly Patterns

```{r hourly_patterns}
# Average trips by hour and day type
hourly_patterns <- indego %>%
  group_by(hour, weekend) %>%
  summarize(avg_trips = n() / n_distinct(date)) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(hourly_patterns, aes(x = hour, y = avg_trips, color = day_type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Average Hourly Ridership Patterns",
    subtitle = "Clear commute patterns on weekdays",
    x = "Hour of Day",
    y = "Average Trips per Hour",
    color = "Day Type"
  ) +
  plotTheme

```

Weekdays show two sharp peaks at hour 8 (morning rush, \~210 trips) and hour 17 (evening rush, \~270 trips), with a notable dip in between, reflecting classic commuter behavior. Weekends show a very different pattern with no morning spike, instead rising gradually from mid-morning and plateauing broadly between hours 12 and 17 at around 150 trips, suggesting recreational rather than commute-driven usage. Overall, weekdays have higher peak demand but weekends sustain more consistent midday ridership.

#### Top Stations

```{r top_stations}
# Most popular origin stations
top_stations <- indego %>%
  count(start_station, start_lat, start_lon, name = "trips") %>%
  arrange(desc(trips)) %>%
  head(20)

kable(top_stations, 
      caption = "Top 20 Indego Stations by Trip Origins",
      format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Load Philadelphia Census Data

```{r load_census}
#| message: false
#| warning: false
#| results: hide

# Get Philadelphia census tracts
philly_census <- get_acs(
  geography = "tract",
  variables = c(
    "B01003_001",  # Total population
    "B19013_001",  # Median household income
    "B08301_001",  # Total commuters
    "B08301_010",  # Commute by transit
    "B02001_002",  # White alone
    "B25077_001"   # Median home value
  ),
  state = "PA",
  county = "Philadelphia",
  year = 2022,
  geometry = TRUE,
  output = "wide"
) %>%
  rename(
    Total_Pop = B01003_001E,
    Med_Inc = B19013_001E,
    Total_Commuters = B08301_001E,
    Transit_Commuters = B08301_010E,
    White_Pop = B02001_002E,
    Med_Home_Value = B25077_001E
  ) %>%
  mutate(
    Percent_Taking_Transit = (Transit_Commuters / Total_Commuters) * 100,
    Percent_White = (White_Pop / Total_Pop) * 100
  ) %>%
  st_transform(crs = 4326)  #transforming to WGS84 for lat/lon matching, currently in NAD83

```

#### Map Philadelphia Context

```{r map_philly}
# Map median income
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = NA) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Context for understanding bike share demand patterns"
  ) +
  # Stations 
  geom_point(
    data = indego,
    aes(x = start_lon, y = start_lat),
    color = "red", size = 0.25, alpha = 0.6
  ) +
  mapTheme
```

#### Join Census Data to Stations

```{r join_census_to_stations}
# Create sf object for stations
stations_sf <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326)

# Spatial join to get census tract for each station
stations_census <- st_join(stations_sf, philly_census, left = TRUE) %>%
  st_drop_geometry()

# Look at the result - investigate whether all of the stations joined to census data -- according to the map above there are stations in non-residential tracts.

stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Add back to trip data
indego_census <- indego %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )


# Prepare data for visualization
stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Create the map showing problem stations
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = "white", size = 0.1) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar,
    na.value = "grey90"
  ) +
  # Stations with census data (small grey dots)
  geom_point(
    data = stations_for_map %>% filter(has_census),
    aes(x = start_lon, y = start_lat),
    color = "grey30", size = 1, alpha = 0.6
  ) +
  # Stations WITHOUT census data (red X marks the spot)
  geom_point(
    data = stations_for_map %>% filter(!has_census),
    aes(x = start_lon, y = start_lat),
    color = "red", size = 1, shape = 4, stroke = 1.5
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Indego stations shown (RED = no census data match)",
    caption = "Red X marks indicate stations that didn't join to census tracts"
  ) +
  mapTheme



```

#### Dealing with missing data

```{r}
# Identify which stations to keep
valid_stations <- stations_census %>%
  filter(!is.na(Med_Inc)) %>%
  pull(start_station)

# Filter trip data to valid stations only
indego_census <- indego %>%
  filter(start_station %in% valid_stations) %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )

```

### Get Weather Data

```{r get_weather}

library(httr2)

weather_raw <- read_csv(
  paste0(
    "https://mesonet.agron.iastate.edu/cgi-bin/request/asos.py?",
    "station=PHL&data=tmpf&data=p01i&data=sknt",
    "&year1=2025&month1=1&day1=1",
    "&year2=2025&month2=3&day2=31",
    "&tz=UTC&format=onlycomma&latlon=no&direct=no"
  ),
  na = c("", "NA", "M", "T")  # M = missing, T = trace precip in ASOS data
)

weather_processed <- weather_raw %>%
  filter(!is.na(valid)) %>%
  mutate(
    interval60    = floor_date(ymd_hms(valid), unit = "hour"),
    Temperature   = as.numeric(tmpf),
    Precipitation = as.numeric(p01i),
    Precipitation = replace_na(Precipitation, 0),
    Wind_Speed    = as.numeric(sknt)
  ) %>%
  filter(!is.na(interval60)) %>%
  group_by(interval60) %>%
  summarize(
    Temperature   = mean(Temperature, na.rm = TRUE),
    Precipitation = sum(Precipitation, na.rm = TRUE),
    Wind_Speed    = mean(Wind_Speed, na.rm = TRUE)
  ) %>%
  mutate(
    Temperature = ifelse(is.nan(Temperature), NA, Temperature),
    Wind_Speed  = ifelse(is.nan(Wind_Speed),  NA, Wind_Speed)
  )

weather_complete <- weather_processed %>%
  complete(interval60 = seq(min(interval60), max(interval60), by = "hour")) %>%
  fill(Temperature, Precipitation, Wind_Speed, .direction = "down")

summary(weather_complete %>% select(Temperature, Precipitation, Wind_Speed))

```

#### Visualize Weather Patterns

```{r visualize-weather}
#| message: false
#| warning: false

ggplot(weather_complete, aes(x = interval60, y = Temperature)) +
  geom_line(color = "#3182bd", alpha = 0.7) +
  geom_smooth(se = FALSE, color = "red") +
  labs(
    title = "Philadelphia Temperature - Q1 2025",
    subtitle = "Winter to early spring transition",
    x = "Date",
    y = "Temperature (°F)"
  ) +
  plotTheme
```

### Aggregate Trips to Station-Hour Level

```{r aggregate-trips}
#| message: false
#| warning: false
#| results: hide

# Count trips by station-hour
trips_panel <- indego_census %>%
  group_by(interval60, start_station, start_lat, start_lon,
           Med_Inc, Percent_Taking_Transit, Percent_White, Total_Pop) %>%
  summarize(Trip_Count = n()) %>%
  ungroup()

# How many station-hour observations?
nrow(trips_panel)

# How many unique stations?
length(unique(trips_panel$start_station))

# How many unique hours?
length(unique(trips_panel$interval60))

hist(trips_panel$Trip_Count)
```

### Create Complete Panel Structure

Not every station has trips every hour. We need a **complete panel** where every station-hour combination exists (even if Trip_Count = 0).

```{r complete-panel}
# Calculate expected panel size

n_stations <- length(unique(trips_panel$start_station))
n_hours <- length(unique(trips_panel$interval60))
expected_rows <- n_stations * n_hours

cat("Expected panel rows:", format(expected_rows, big.mark = ","), "\n")
cat("Current rows:", format(nrow(trips_panel), big.mark = ","), "\n")
cat("Missing rows:", format(expected_rows - nrow(trips_panel), big.mark = ","), "\n")

# Create complete panel
study_panel <- expand_grid(
  interval60 = unique(trips_panel$interval60),
  start_station = unique(trips_panel$start_station)
) %>%
  # Join trip counts
  left_join(trips_panel %>% 
              select(interval60, start_station, Trip_Count),
            by = c("interval60", "start_station")) %>%
  # Replace NA trip counts with 0
  mutate(Trip_Count = replace_na(Trip_Count, 0))

glimpse(study_panel)
hist(study_panel$Trip_Count)

# Fill in station attributes (they're the same for all hours)
station_attributes <- trips_panel %>%
  group_by(start_station) %>%
  summarize(
    start_lat = first(start_lat),
    start_lon = first(start_lon),
    Med_Inc = first(Med_Inc),
    Percent_Taking_Transit = first(Percent_Taking_Transit),
    Percent_White = first(Percent_White),
    Total_Pop = first(Total_Pop)
  )

glimpse(station_attributes)

study_panel <- study_panel %>%
  left_join(station_attributes, by = "start_station")

glimpse(study_panel)

# Verify we have complete panel
cat("Complete panel rows:", format(nrow(study_panel), big.mark = ","), "\n")
```

### Add Time Features

```{r add_time_features}
study_panel <- study_panel %>%
  mutate(
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

glimpse(study_panel)
```

### Join Weather Data

```{r join_weather}
study_panel <- study_panel %>%
  left_join(weather_complete, by = "interval60")

glimpse(study_panel)

# Check for missing values
summary(study_panel %>% select(Trip_Count, Temperature, Precipitation))
```

### CREATE TEMPORAL LAG VARIABLES

```{r create_lags}
# Sort by station and time
study_panel <- study_panel %>%
  arrange(start_station, interval60)

# Create lag variables WITHIN each station
study_panel <- study_panel %>%
  group_by(start_station) %>%
  mutate(
    lag1Hour = lag(Trip_Count, 1),
    lag2Hours = lag(Trip_Count, 2),
    lag3Hours = lag(Trip_Count, 3),
    lag12Hours = lag(Trip_Count, 12),
    lag1day = lag(Trip_Count, 24)
  ) %>%
  ungroup()

# Remove rows with NA lags (first 24 hours for each station)
study_panel_complete <- study_panel %>%
  filter(!is.na(lag1day))

cat("Rows after removing NA lags:", format(nrow(study_panel_complete), big.mark = ","), "\n")
```

### Visualize Lag Correlations

```{r lag_correlations}
# Sample one station to visualize
example_station <- study_panel_complete %>%
  filter(start_station == first(start_station)) %>%
  head(168)  # One week

# Plot actual vs lagged demand
ggplot(example_station, aes(x = interval60)) +
  geom_line(aes(y = Trip_Count, color = "Current"), linewidth = 1) +
  geom_line(aes(y = lag1Hour, color = "1 Hour Ago"), linewidth = 1, alpha = 0.7) +
  geom_line(aes(y = lag1day, color = "24 Hours Ago"), linewidth = 1, alpha = 0.7) +
  scale_color_manual(values = c(
    "Current" = "#08519c",
    "1 Hour Ago" = "#3182bd",
    "24 Hours Ago" = "#6baed6"
  )) +
  labs(
    title = "Temporal Lag Patterns at One Station",
    subtitle = "Past demand predicts future demand",
    x = "Date-Time",
    y = "Trip Count",
    color = "Time Period"
  ) +
  plotTheme
```

### TEMPORAL TRAIN/TEST SPLIT



```{r temporal-split}
# Split by week
# Q1 has weeks 1-13 (Jan-Mar)
# Train on weeks 1-9 (Jan 1 - early March)
# Test on weeks 10-13 (rest of March)

# Which stations have trips in BOTH early and late periods?
early_stations <- study_panel_complete %>%
  filter(week < 10) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

late_stations <- study_panel_complete %>%
  filter(week >= 10) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

# Keep only stations that appear in BOTH periods
common_stations <- intersect(early_stations, late_stations)


# Filter panel to only common stations
study_panel_complete <- study_panel_complete %>%
  filter(start_station %in% common_stations)

# NOW create train/test split
train <- study_panel_complete %>%
  filter(week < 10)

test <- study_panel_complete %>%
  filter(week >= 10)

cat("Training observations:", format(nrow(train), big.mark = ","), "\n")
cat("Testing observations:", format(nrow(test), big.mark = ","), "\n")
cat("Training date range:", min(train$date), "to", max(train$date), "\n")
cat("Testing date range:", min(test$date), "to", max(test$date), "\n")

```

### BUILD PREDICTIVE MODELS

We'll build 5 models with increasing complexity to see what improves predictions.

#### Model 1: Baseline (Time + Weather)

```{r model1}
#| results: hide
# Create day of week factor with treatment (dummy) coding
train <- train %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(train$dotw_simple) <- contr.treatment(7)

# Now run the model
model1 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation,
  data = train
)

summary(model1)
```

The model uses Monday as the baseline. Each coefficient represents the difference in expected trips per station-hour compared to Monday - dow_simple2 = Tuesday..

**Weekday Pattern (Tue-Fri):**

-   All weekdays have positive coefficients (0.029 to 0.052)
-   Tuesday has the highest weekday effect (+0.052)
-   Weekdays likely benefit from concentrated commuting patterns

**Weekend Pattern (Sat-Sun):**

-   Both weekend days have negative coefficients (-0.061 and -0.065)
-   This means FEWER trips per station-hour than Monday

**Hourly Interpretation**

Hour Coefficient Interpretation 0 (baseline) 0.000 trips/hour (midnight) 1 -0.018 slightly fewer than midnight ... 6 +0.151 morning activity starting 7 +0.276 morning rush building 8 +0.487 PEAK morning rush 9 +0.350 post-rush ... 17 +0.568 PEAK evening rush (5 PM!) 18 +0.389 evening declining ... 23 +0.034 late night minimal


#### Model 2: Add Temporal Lags

```{r model2}
#| results: hide
model2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day,
  data = train
)

summary(model2)
```


#### Model 3: Add Demographics

```{r model3}
#| results: hide
model3 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White,
  data = train
)

summary(model3)
```

#### Model 4: Add Station Fixed Effects

```{r model4}
#| results: hide
model4 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station),
  data = train
)

# Summary too long with all station dummies, just show key metrics
cat("Model 4 R-squared:", summary(model4)$r.squared, "\n")
cat("Model 4 Adj R-squared:", summary(model4)$adj.r.squared, "\n")
```


#### Model 5: Add Rush Hour Interaction

```{r model5}
#| results: hide
model5 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day + rush_hour + as.factor(month) +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station) +
    rush_hour * weekend,  # Rush hour effects different on weekends
  data = train
)

cat("Model 5 R-squared:", summary(model5)$r.squared, "\n")
cat("Model 5 Adj R-squared:", summary(model5)$adj.r.squared, "\n")
```

### MODEL EVALUATION

#### Calculate Predictions and MAE

```{r calculate_mae}
# Get predictions on test set

# Create day of week factor with treatment (dummy) coding
test <- test %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(test$dotw_simple) <- contr.treatment(7)

test <- test %>%
  mutate(
    pred1 = predict(model1, newdata = test),
    pred2 = predict(model2, newdata = test),
    pred3 = predict(model3, newdata = test),
    pred4 = predict(model4, newdata = test),
    pred5 = predict(model5, newdata = test)
  )

# Calculate MAE for each model
mae_results <- data.frame(
  Model = c(
    "1. Time + Weather",
    "2. + Temporal Lags",
    "3. + Demographics",
    "4. + Station FE",
    "5. + Rush Hour Interaction"
  ),
  MAE = c(
    mean(abs(test$Trip_Count - test$pred1), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred2), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred3), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred4), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred5), na.rm = TRUE)
  )
)

kable(mae_results, 
      digits = 2,
      caption = "Mean Absolute Error by Model (Test Set)",
      col.names = c("Model", "MAE (trips)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

mae_results_q1 <- mae_results
```

#### Visualize Model Comparison

```{r compare_models}
ggplot(mae_results, aes(x = reorder(Model, -MAE), y = MAE)) +
  geom_col(fill = "#3182bd", alpha = 0.8) +
  geom_text(aes(label = round(MAE, 2)), vjust = -0.5) +
  labs(
    title = "Model Performance Comparison",
    subtitle = "Lower MAE = Better Predictions",
    x = "Model",
    y = "Mean Absolute Error (trips)"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


### SPACE-TIME ERROR ANALYSIS

#### Observed vs. Predicted

Let's use our best model (Model 2) for error analysis.

```{r obs_vs_pred}
#| message: false
#| warning: false

test <- test %>%
  mutate(
    error = Trip_Count - pred2,
    abs_error = abs(error),
    time_of_day = case_when(
      hour < 7 ~ "Overnight",
      hour >= 7 & hour < 10 ~ "AM Rush",
      hour >= 10 & hour < 15 ~ "Mid-Day",
      hour >= 15 & hour <= 18 ~ "PM Rush",
      hour > 18 ~ "Evening"
    )
  )

# Scatter plot by time and day type
ggplot(test, aes(x = Trip_Count, y = pred2)) +
  geom_point(alpha = 0.2, color = "#3182bd") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  facet_grid(weekend ~ time_of_day) +
  labs(
    title = "Observed vs. Predicted Bike Trips",
    subtitle = "Model 2 performance by time period",
    x = "Observed Trips",
    y = "Predicted Trips",
    caption = "Red line = perfect predictions; Green line = actual model fit"
  ) +
  plotTheme
```



#### Spatial Error Patterns


```{r spatial_errors}
# Calculate MAE by station
station_errors <- test %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

## Create Two Maps Side-by-Side with Proper Legends (sorry these maps are ugly)

# Calculate station errors
station_errors <- test %>%
  filter(!is.na(pred2)) %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs(Trip_Count - pred2), na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE\n(trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),  # Fewer, cleaner breaks
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors",
       subtitle = "Higher in Center City") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand
p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg\nDemand",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),  # Clear breaks
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand",
       subtitle = "Trips per station-hour") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))




# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE (trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand  
p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg Demand (trips/hour)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Combine
(p1 | p2)

```


#### Temporal Error Patterns

When are we most wrong?

```{r temporal_errors}
# MAE by time of day and day type
temporal_errors <- test %>%
  group_by(time_of_day, weekend) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(temporal_errors, aes(x = time_of_day, y = MAE, fill = day_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Prediction Errors by Time Period",
    subtitle = "When is the model struggling most?",
    x = "Time of Day",
    y = "Mean Absolute Error (trips)",
    fill = "Day Type"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Errors and Demographics

Are prediction errors related to neighborhood characteristics?

```{r errors_demographics}
#| message: false
#| warning: false

# Join demographic data to station errors
station_errors_demo <- station_errors %>%
  left_join(
    station_attributes %>% select(start_station, Med_Inc, Percent_Taking_Transit, Percent_White),
    by = "start_station"
  ) %>%
  filter(!is.na(Med_Inc))

# Create plots
p1 <- ggplot(station_errors_demo, aes(x = Med_Inc, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Errors vs. Median Income", x = "Median Income", y = "MAE") +
  plotTheme

p2 <- ggplot(station_errors_demo, aes(x = Percent_Taking_Transit, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Transit Usage", x = "% Taking Transit", y = "MAE") +
  plotTheme

p3 <- ggplot(station_errors_demo, aes(x = Percent_White, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Race", x = "% White", y = "MAE") +
  plotTheme

(p1 | p2) / (p3 | plot_spacer())
```



## 4.2 Q2 2025 Analysis

### Load and Process Indego Trip Data (Q2 2025)
```{r}
indego_q2 <- read_csv(here("data/indego-trips-2025-q2.csv"))
cat("Total trips in Q2 2025:", nrow(indego_q2), "\n")
cat("Date range:", min(mdy_hm(indego_q2$start_time)), "to", max(mdy_hm(indego_q2$start_time)), "\n")
cat("Unique start stations:", length(unique(indego_q2$start_station)), "\n")
table(indego_q2$trip_route_category)
table(indego_q2$passholder_type)
table(indego_q2$bike_type)
```
```{r create-time-bins-q2}
indego_q2 <- indego_q2 %>%
  mutate(
    start_datetime = mdy_hm(start_time),
    end_datetime   = mdy_hm(end_time),
    interval60 = floor_date(start_datetime, unit = "hour"),
    week    = week(interval60),
    month   = month(interval60, label = TRUE),
    dotw    = wday(interval60, label = TRUE),
    hour    = hour(interval60),
    date    = as.Date(interval60),
    weekend   = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )
head(indego_q2 %>% select(start_datetime, interval60, week, dotw, hour, weekend))
```

#### Daily Trip
```{r trips_over_time_q2}
#| message: false
#| warning: false
daily_trips_q2 <- indego_q2 %>%
  group_by(date) %>%
  summarize(trips = n())

ggplot(daily_trips_q2, aes(x = date, y = trips)) +
  geom_line(color = "#3182bd", linewidth = 1) +
  geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
  labs(
    title = "Indego Daily Ridership - Q2 2025",
    subtitle = "Spring to early summer demand patterns in Philadelphia",
    x = "Date",
    y = "Daily Trips",
    caption = "Source: Indego bike share"
  ) +
  plotTheme
```
It has a different pattern from Q1, that there is no spikes.


#### Hourly Patterns
```{r hourly_patterns_q2}
hourly_patterns_q2 <- indego_q2 %>%
  group_by(hour, weekend) %>%
  summarize(avg_trips = n() / n_distinct(date)) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(hourly_patterns_q2, aes(x = hour, y = avg_trips, color = day_type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Average Hourly Ridership Patterns - Q2 2025",
    subtitle = "Clear commute patterns on weekdays",
    x = "Hour of Day",
    y = "Average Trips per Hour",
    color = "Day Type"
  ) +
  plotTheme
```
#### Top Stations
```{r top_stations_q2}
top_stations_q2 <- indego_q2 %>%
  count(start_station, start_lat, start_lon, name = "trips") %>%
  arrange(desc(trips)) %>%
  head(20)

kable(top_stations_q2,
      caption = "Top 20 Indego Stations by Trip Origins - Q2 2025",
      format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Join Census Data to Q2 Stations
`philly_census` is already loaded from Q1. We reuse it and rejoin to Q2 stations.
```{r join_census_q2}
#| message: false
#| warning: false
stations_sf_q2 <- indego_q2 %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326)

stations_census_q2 <- st_join(stations_sf_q2, philly_census, left = TRUE) %>%
  st_drop_geometry()

stations_for_map_q2 <- indego_q2 %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(stations_census_q2 %>% select(start_station, Med_Inc), by = "start_station") %>%
  mutate(has_census = !is.na(Med_Inc))

ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = "white", size = 0.1) +
  scale_fill_viridis(option = "viridis", name = "Median\nIncome",
                     labels = scales::dollar, na.value = "grey90") +
  geom_point(data = stations_for_map_q2 %>% filter(has_census),
             aes(x = start_lon, y = start_lat), color = "grey30", size = 1, alpha = 0.6) +
  geom_point(data = stations_for_map_q2 %>% filter(!has_census),
             aes(x = start_lon, y = start_lat), color = "red", size = 1, shape = 4, stroke = 1.5) +
  labs(title = "Philadelphia Median Household Income by Census Tract",
       subtitle = "Indego Q2 stations shown (RED = no census data match)",
       caption = "Red X marks indicate stations that did not join to census tracts") +
  mapTheme
```
```{r missing_data_q2}
valid_stations_q2 <- stations_census_q2 %>%
  filter(!is.na(Med_Inc)) %>%
  pull(start_station)

indego_census_q2 <- indego_q2 %>%
  filter(start_station %in% valid_stations_q2) %>%
  left_join(
    stations_census_q2 %>%
      select(start_station, Med_Inc, Percent_Taking_Transit, Percent_White, Total_Pop),
    by = "start_station"
  )
```
### Get weather Data for Q2 2025
```{r get_weather_q2}
weather_raw_q2 <- read_csv(
  paste0(
    "https://mesonet.agron.iastate.edu/cgi-bin/request/asos.py?",
    "station=PHL&data=tmpf&data=p01i&data=sknt",
    "&year1=2025&month1=4&day1=1",
    "&year2=2025&month2=6&day2=30",
    "&tz=UTC&format=onlycomma&latlon=no&direct=no"
  ),
  na = c("", "NA", "M", "T")
)

weather_processed_q2 <- weather_raw_q2 %>%
  filter(!is.na(valid)) %>%
  mutate(
    interval60    = floor_date(ymd_hms(valid), unit = "hour"),
    Temperature   = as.numeric(tmpf),
    Precipitation = as.numeric(p01i),
    Precipitation = replace_na(Precipitation, 0),
    Wind_Speed    = as.numeric(sknt)
  ) %>%
  filter(!is.na(interval60)) %>%
  group_by(interval60) %>%
  summarize(
    Temperature   = mean(Temperature, na.rm = TRUE),
    Precipitation = sum(Precipitation, na.rm = TRUE),
    Wind_Speed    = mean(Wind_Speed, na.rm = TRUE)
  ) %>%
  mutate(
    Temperature = ifelse(is.nan(Temperature), NA, Temperature),
    Wind_Speed  = ifelse(is.nan(Wind_Speed),  NA, Wind_Speed)
  )

weather_complete_q2 <- weather_processed_q2 %>%
  complete(interval60 = seq(min(interval60), max(interval60), by = "hour")) %>%
  fill(Temperature, Precipitation, Wind_Speed, .direction = "down")

summary(weather_complete_q2 %>% select(Temperature, Precipitation, Wind_Speed))
```

```{r visualize-weather-q2}
#| message: false
#| warning: false
ggplot(weather_complete_q2, aes(x = interval60, y = Temperature)) +
  geom_line(color = "#3182bd", alpha = 0.7) +
  geom_smooth(se = FALSE, color = "red") +
  labs(
    title = "Philadelphia Temperature - Q2 2025",
    subtitle = "Spring to early summer transition",
    x = "Date",
    y = "Temperature (F)"
  ) +
  plotTheme
```

### Create Space-Time Panel
```{r aggregate-trips-q2}
#| message: false
#| warning: false
trips_panel_q2 <- indego_census_q2 %>%
  group_by(interval60, start_station, start_lat, start_lon,
           Med_Inc, Percent_Taking_Transit, Percent_White, Total_Pop) %>%
  summarize(Trip_Count = n()) %>%
  ungroup()

nrow(trips_panel_q2)
length(unique(trips_panel_q2$start_station))
length(unique(trips_panel_q2$interval60))
hist(trips_panel_q2$Trip_Count)
```
```{r complete-panel-q2}
n_stations_q2 <- length(unique(trips_panel_q2$start_station))
n_hours_q2    <- length(unique(trips_panel_q2$interval60))
expected_rows_q2 <- n_stations_q2 * n_hours_q2

cat("Expected panel rows:", format(expected_rows_q2, big.mark = ","), "\n")
cat("Current rows:", format(nrow(trips_panel_q2), big.mark = ","), "\n")
cat("Missing rows:", format(expected_rows_q2 - nrow(trips_panel_q2), big.mark = ","), "\n")

study_panel_q2 <- expand_grid(
  interval60    = unique(trips_panel_q2$interval60),
  start_station = unique(trips_panel_q2$start_station)
) %>%
  left_join(trips_panel_q2 %>% select(interval60, start_station, Trip_Count),
            by = c("interval60", "start_station")) %>%
  mutate(Trip_Count = replace_na(Trip_Count, 0))

station_attributes_q2 <- trips_panel_q2 %>%
  group_by(start_station) %>%
  summarize(
    start_lat = first(start_lat),
    start_lon = first(start_lon),
    Med_Inc = first(Med_Inc),
    Percent_Taking_Transit = first(Percent_Taking_Transit),
    Percent_White = first(Percent_White),
    Total_Pop = first(Total_Pop)
  )

study_panel_q2 <- study_panel_q2 %>%
  left_join(station_attributes_q2, by = "start_station")

cat("Complete panel rows:", format(nrow(study_panel_q2), big.mark = ","), "\n")
```
```{r add_time_features_q2}
study_panel_q2 <- study_panel_q2 %>%
  mutate(
    week      = week(interval60),
    month     = month(interval60, label = TRUE),
    dotw      = wday(interval60, label = TRUE),
    hour      = hour(interval60),
    date      = as.Date(interval60),
    weekend   = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )
```

```{r join_weather_q2}
study_panel_q2 <- study_panel_q2 %>%
  left_join(weather_complete_q2, by = "interval60")

summary(study_panel_q2 %>% select(Trip_Count, Temperature, Precipitation))
```
#### Create Temporal Lag Variables
```{r create_lags_q2}
study_panel_q2 <- study_panel_q2 %>%
  arrange(start_station, interval60) %>%
  group_by(start_station) %>%
  mutate(
    lag1Hour  = lag(Trip_Count, 1),
    lag2Hours = lag(Trip_Count, 2),
    lag3Hours = lag(Trip_Count, 3),
    lag12Hours = lag(Trip_Count, 12),
    lag1day   = lag(Trip_Count, 24)
  ) %>%
  ungroup()

study_panel_complete_q2 <- study_panel_q2 %>%
  filter(!is.na(lag1day))

cat("Rows after removing NA lags:", format(nrow(study_panel_complete_q2), big.mark = ","), "\n")
```
```{r lag_correlations_q2}
example_station_q2 <- study_panel_complete_q2 %>%
  filter(start_station == first(start_station)) %>%
  head(168)

ggplot(example_station_q2, aes(x = interval60)) +
  geom_line(aes(y = Trip_Count, color = "Current"), linewidth = 1) +
  geom_line(aes(y = lag1Hour,  color = "1 Hour Ago"), linewidth = 1, alpha = 0.7) +
  geom_line(aes(y = lag1day,   color = "24 Hours Ago"), linewidth = 1, alpha = 0.7) +
  scale_color_manual(values = c("Current" = "#08519c", "1 Hour Ago" = "#3182bd", "24 Hours Ago" = "#6baed6")) +
  labs(title = "Temporal Lag Patterns at One Station - Q2 2025",
       subtitle = "Past demand predicts future demand",
       x = "Date-Time", y = "Trip Count", color = "Time Period") +
  plotTheme
```

### Build Predictive models

#### Temporal train and test split
Q2 spans weeks 14 to 26. We train on weeks 14-22 (April through May) and test on weeks 23-26 (June).

```{r temporal-split-q2}
early_stations_q2 <- study_panel_complete_q2 %>%
  filter(week < 23, Trip_Count > 0) %>%
  distinct(start_station) %>% pull(start_station)

late_stations_q2 <- study_panel_complete_q2 %>%
  filter(week >= 23, Trip_Count > 0) %>%
  distinct(start_station) %>% pull(start_station)

common_stations_q2 <- intersect(early_stations_q2, late_stations_q2)

study_panel_complete_q2 <- study_panel_complete_q2 %>%
  filter(start_station %in% common_stations_q2)

train_q2 <- study_panel_complete_q2 %>% filter(week < 23)
test_q2  <- study_panel_complete_q2 %>% filter(week >= 23)

cat("Training observations:", format(nrow(train_q2), big.mark = ","), "\n")
cat("Testing observations:",  format(nrow(test_q2),  big.mark = ","), "\n")
cat("Training date range:", min(train_q2$date), "to", max(train_q2$date), "\n")
cat("Testing date range:",  min(test_q2$date),  "to", max(test_q2$date), "\n")
```


#### Model 1: Baseline (Time + Weather)

```{r model1_q2}
#| results: hide
train_q2 <- train_q2 %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon","Tue","Wed","Thu","Fri","Sat","Sun")))
contrasts(train_q2$dotw_simple) <- contr.treatment(7)

model1_q2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation,
  data = train_q2
)
summary(model1_q2)
```
#### Model 2: Add Temporal Lags
```{r model2_q2}
#| results: hide
model2_q2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day,
  data = train_q2
)
summary(model2_q2)
```
#### Model 3: Add Demographics
```{r model3_q2}
#| results: hide
model3_q2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White,
  data = train_q2
)
summary(model3_q2)
```
#### Model 4: Add Station Fixed Effects
```{r model4_q2}
#| results: hide
model4_q2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station),
  data = train_q2
)
cat("Model 4 R-squared:", summary(model4_q2)$r.squared, "\n")
cat("Model 4 Adj R-squared:", summary(model4_q2)$adj.r.squared, "\n")
```
#### Model 5: Add Rush Hour Interaction
```{r model5_q2}
#| results: hide
model5_q2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day + rush_hour + as.factor(month) +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station) +
    rush_hour * weekend,
  data = train_q2
)
cat("Model 5 R-squared:", summary(model5_q2)$r.squared, "\n")
cat("Model 5 Adj R-squared:", summary(model5_q2)$adj.r.squared, "\n")
```
#### Model Evaluation
```{r calculate_mae_q2}
test_q2 <- test_q2 %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon","Tue","Wed","Thu","Fri","Sat","Sun")))
contrasts(test_q2$dotw_simple) <- contr.treatment(7)

test_q2 <- test_q2 %>%
  mutate(
    pred1 = predict(model1_q2, newdata = test_q2),
    pred2 = predict(model2_q2, newdata = test_q2),
    pred3 = predict(model3_q2, newdata = test_q2),
    pred4 = predict(model4_q2, newdata = test_q2),
    pred5 = predict(model5_q2, newdata = test_q2)
  )

mae_results_q2 <- data.frame(
  Model = c("1. Time + Weather", "2. + Temporal Lags", "3. + Demographics",
            "4. + Station FE", "5. + Rush Hour Interaction"),
  MAE = c(
    mean(abs(test_q2$Trip_Count - test_q2$pred1), na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred2), na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred3), na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred4), na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred5), na.rm = TRUE)
  ),
  RMSE = c(
    sqrt(mean((test_q2$Trip_Count - test_q2$pred1)^2, na.rm = TRUE)),
    sqrt(mean((test_q2$Trip_Count - test_q2$pred2)^2, na.rm = TRUE)),
    sqrt(mean((test_q2$Trip_Count - test_q2$pred3)^2, na.rm = TRUE)),
    sqrt(mean((test_q2$Trip_Count - test_q2$pred4)^2, na.rm = TRUE)),
    sqrt(mean((test_q2$Trip_Count - test_q2$pred5)^2, na.rm = TRUE))
  ),
  R2 = c(
    summary(model1_q2)$r.squared,
    summary(model2_q2)$r.squared,
    summary(model3_q2)$r.squared,
    summary(model4_q2)$r.squared,
    summary(model5_q2)$r.squared
  )
)

kable(mae_results_q2, digits = 3,
      caption = "Model Performance - Q2 2025 (Test Set)",
      col.names = c("Model", "MAE (trips)", "RMSE (trips)", "R²")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```
**Interpretation**: The largest improvement comes from adding temporal lags in Model 2, reducing MAE from 0.835 to 0.690 and improving R² from 0.112 to 0.305, confirming that past demand is by far the strongest predictor. Adding demographics in Model 3 yields only marginal gains, while station fixed effects in Model 4 provide a small but meaningful improvement in R² (0.337) and RMSE (1.040), capturing baseline differences in busyness across stations. Interestingly, Model 5 slightly increases MAE and RMSE despite a higher R², suggesting the rush hour interaction may be overfitting to the training data and not generalizing well to the June test period.

#### Visualize q2 model comparison

```{r compare_models_q2}
ggplot(mae_results_q2, aes(x = reorder(Model, -MAE), y = MAE)) +
  geom_col(fill = "#3182bd", alpha = 0.8) +
  geom_text(aes(label = round(MAE, 2)), vjust = -0.5) +
  labs(title = "Model Performance Comparison - Q2 2025",
       subtitle = "Lower MAE = Better Predictions",
       x = "Model", y = "Mean Absolute Error (trips)") +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Q1 vs Q2 Model Comparison

```{r compare_quarters}
comparison <- bind_rows(
  mae_results_q1 %>% mutate(Quarter = "Q1 2025"),
  mae_results_q2 %>% mutate(Quarter = "Q2 2025")
)

ggplot(comparison, aes(x = Model, y = MAE, fill = Quarter)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Q1 2025" = "#08519c", "Q2 2025" = "#6baed6")) +
  labs(title = "MAE Comparison: Q1 vs Q2 2025",
       subtitle = "Does the model perform better in spring than winter?",
       x = "Model", y = "Mean Absolute Error (trips)", fill = "Quarter") +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
**Interpretation**:
Q1 consistently outperforms Q2 across all five models, with Q1 MAE ranging from 0.63 to 0.53 compared to Q2's 0.84 to 0.71. This difference likely reflects the higher and more variable demand in spring and early summer, where warmer weather, recreational riding, and special events introduce greater unpredictability that the model struggles to capture. Both quarters show the same pattern of temporal lags providing the biggest single improvement, confirming that past demand is the most important predictor regardless of season. In Q2, additional complexity from demographics and station fixed effects yields smaller marginal gains, suggesting that Q2 demand is driven more by weather and temporal factors than by neighborhood characteristics.

### Space-Time Error Analysis

#### Observed vs. Predicted
```{r obs_vs_pred_q2}
#| message: false
#| warning: false
test_q2 <- test_q2 %>%
  mutate(
    error = Trip_Count - pred2,
    abs_error = abs(error),
    time_of_day = case_when(
      hour < 7 ~ "Overnight",
      hour >= 7  & hour < 10 ~ "AM Rush",
      hour >= 10 & hour < 15 ~ "Mid-Day",
      hour >= 15 & hour <= 18 ~ "PM Rush",
      hour > 18 ~ "Evening"
    )
  )

p1 <- ggplot(test, aes(x = Trip_Count, y = pred2)) +
  geom_point(alpha = 0.2, color = "#08519c") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  facet_grid(weekend ~ time_of_day) +
  labs(title = "Q1 2025: Observed vs. Predicted",
       x = "Observed Trips", y = "Predicted Trips",
       caption = "Red line = perfect predictions; Green line = actual model fit") +
  plotTheme

p2 <- ggplot(test_q2, aes(x = Trip_Count, y = pred2)) +
  geom_point(alpha = 0.2, color = "#6baed6") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  facet_grid(weekend ~ time_of_day) +
  labs(title = "Q2 2025: Observed vs. Predicted",
       x = "Observed Trips", y = "Predicted Trips",
       caption = "Red line = perfect predictions; Green line = actual model fit") +
  plotTheme

p1 / p2
```

#### Spatial Error Patterns
```{r spatial_errors_q2}
station_errors_q2 <- test_q2 %>%
  filter(!is.na(pred2)) %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs(Trip_Count - pred2), na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(data = station_errors, aes(x = start_lon, y = start_lat, color = MAE),
             size = 3.5, alpha = 0.7) +
  scale_color_viridis(option = "plasma", name = "MAE (trips)", direction = -1,
                      breaks = c(0.5, 1.0, 1.5), labels = c("0.5", "1.0", "1.5")) +
  labs(title = "Q1 2025: Prediction Errors") +
  mapTheme +
  theme(legend.position = "bottom",
        legend.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 9),
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +
  guides(color = guide_colorbar(barwidth = 12, barheight = 1,
                                title.position = "top", title.hjust = 0.5))

p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(data = station_errors_q2, aes(x = start_lon, y = start_lat, color = MAE),
             size = 3.5, alpha = 0.7) +
  scale_color_viridis(option = "plasma", name = "MAE (trips)", direction = -1,
                      breaks = c(0.5, 1.0, 1.5), labels = c("0.5", "1.0", "1.5")) +
  labs(title = "Q2 2025: Prediction Errors") +
  mapTheme +
  theme(legend.position = "bottom",
        legend.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 9),
        plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +
  guides(color = guide_colorbar(barwidth = 12, barheight = 1,
                                title.position = "top", title.hjust = 0.5))

(p1 | p2)
```

**Interpretation**:
Similar pattern to Q1. Prediction errors are spatially clustered and strongly correlated with average demand, with the highest MAE concentrated in Center City and the area just north of it where ridership is also highest. This is expected since high-demand stations experience greater absolute variability that the model struggles to capture. Outer neighborhoods with lower demand show lower MAE, but this may reflect a floor effect rather than better model fit. The spatial overlap between high-error and high-demand stations suggests the model systematically underpredicts at the busiest locations, which is the most operationally problematic pattern since those are precisely the stations that need accurate forecasts for effective rebalancing.

#### Temporal Error Patterns
```{r temporal_errors_q2}
temporal_errors_q2 <- test_q2 %>%
  group_by(time_of_day, weekend) %>%
  summarize(MAE = mean(abs_error, na.rm = TRUE), .groups = "drop") %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

temporal_comparison <- bind_rows(
  temporal_errors %>% mutate(Quarter = "Q1 2025"),
  temporal_errors_q2 %>% mutate(Quarter = "Q2 2025")
)

ggplot(temporal_comparison, aes(x = time_of_day, y = MAE, fill = Quarter)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Q1 2025" = "#08519c", "Q2 2025" = "#6baed6")) +
  facet_wrap(~ day_type) +
  labs(title = "Prediction Errors by Time Period: Q1 vs Q2 2025",
       subtitle = "When is the model struggling most?",
       x = "Time of Day", y = "Mean Absolute Error (trips)", fill = "Quarter") +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Interpretation**:
For q2, PM Rush hours show the highest prediction errors for both weekdays (~0.97) and weekends (~0.93), indicating the model struggles most when afternoon demand peaks are at their most intense and variable in Q2. AM Rush and Evening periods also show elevated errors, while Overnight hours have the lowest MAE (~0.30) as expected given minimal and highly predictable demand. Notably, weekends tend to have slightly lower errors than weekdays during peak periods, likely because weekend demand is more gradual and spread out, making it easier to predict. The consistently high errors during rush hours suggest the model is missing features that capture day-to-day variability in commute intensity, such as weather interactions with peak hours or event-driven demand spikes unique to spring.


#### Demographic patterns
```{r errors_demographics_q2}
#| message: false
#| warning: false
station_errors_demo_q2 <- station_errors_q2 %>%
  left_join(
    station_attributes_q2 %>% select(start_station, Med_Inc, Percent_Taking_Transit, Percent_White),
    by = "start_station"
  ) %>%
  filter(!is.na(Med_Inc)) %>%
  mutate(Quarter = "Q2 2025")

station_errors_demo_combined <- station_errors_demo %>%
  mutate(Quarter = "Q1 2025") %>%
  bind_rows(station_errors_demo_q2)

p1 <- ggplot(station_errors_demo_combined, aes(x = Med_Inc, y = MAE, color = Quarter)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c("Q1 2025" = "#08519c", "Q2 2025" = "#6baed6")) +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Errors vs. Median Income", x = "Median Income", y = "MAE") +
  plotTheme

p2 <- ggplot(station_errors_demo_combined, aes(x = Percent_Taking_Transit, y = MAE, color = Quarter)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c("Q1 2025" = "#08519c", "Q2 2025" = "#6baed6")) +
  labs(title = "Errors vs. Transit Usage", x = "% Taking Transit", y = "MAE") +
  plotTheme

p3 <- ggplot(station_errors_demo_combined, aes(x = Percent_White, y = MAE, color = Quarter)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c("Q1 2025" = "#08519c", "Q2 2025" = "#6baed6")) +
  labs(title = "Errors vs. Race", x = "% White", y = "MAE") +
  plotTheme

(p1 | p2) / (p3 | plot_spacer())
```

**Interpretation**: 
For q2, All three plots reveal systematic demographic patterns in prediction errors. Higher-income neighborhoods show higher MAE, likely because wealthier areas tend to be in Center City where demand is more intense and variable. The negative relationship with transit usage suggests that stations in transit-dependent neighborhoods, typically lower-income communities, are actually easier to predict, possibly because their ridership patterns are more routine and less influenced by discretionary spring activity. The positive relationship with percent White mirrors the income pattern, reflecting the spatial overlap between race, wealth, and Center City location in Philadelphia. Together these findings raise equity concerns: the model performs worst in affluent, majority-white neighborhoods where Indego is already well-resourced, while performing better in lower-income, higher-transit communities that may have less margin for rebalancing errors.

#### Feature Engineering and Model Improvement

Three new features are added motivated by Q2-specific conditions: `perfect_weather` flags hours with temperatures between 60 and 75°F and no precipitation, capturing ideal biking conditions unique to spring. `semester_end` marks the final weeks of university spring semesters at Penn, Drexel, and Temple (April 21 to May 15), when student activity patterns shift significantly. `memorial_day` flags May 26th as a federal holiday with distinct recreational demand patterns.

**I choose model 4 as my best model** because it achieves the lowest MAE (0.687) and RMSE (1.040) on the test set among all five models. While Model 5 adds a rush hour and weekend interaction term and achieves a marginally higher R² (0.341 vs 0.337), its MAE and RMSE both increase relative to Model 4, suggesting the interaction term overfits the training data and does not generalize well to the June test period. 

```{r feature_engineering_q2}
train_q2 <- train_q2 %>%
  mutate(
    perfect_weather = ifelse(Temperature >= 60 & Temperature <= 75 & Precipitation == 0, 1, 0),
    semester_end    = ifelse(date >= as.Date("2025-04-21") & date <= as.Date("2025-05-15"), 1, 0),
    memorial_day    = ifelse(date == as.Date("2025-05-26"), 1, 0)
  )

test_q2 <- test_q2 %>%
  mutate(
    perfect_weather = ifelse(Temperature >= 60 & Temperature <= 75 & Precipitation == 0, 1, 0),
    semester_end    = ifelse(date >= as.Date("2025-04-21") & date <= as.Date("2025-05-15"), 1, 0),
    memorial_day    = ifelse(date == as.Date("2025-05-26"), 1, 0)
  )
```

### Model 6: Model 4 + New Features (OLS)

```{r model6_q2}
#| results: hide
model6_q2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station) +
    perfect_weather + semester_end + memorial_day,
  data = train_q2
)
cat("Model 6 R-squared:", summary(model6_q2)$r.squared, "\n")
cat("Model 6 Adj R-squared:", summary(model6_q2)$adj.r.squared, "\n")
```
### Model 7: Model 4 + New Features (Poisson)
```{r model_poisson_q2}
#| warning: false
#| results: hide
library(fastglm)

# Build model matrix for train
X_train <- model.matrix(
  ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station) +
    perfect_weather + semester_end + memorial_day,
  data = train_q2
)

y_train <- train_q2$Trip_Count

# Fit fastglm
model_poisson_q2 <- fastglm(X_train, y_train, family = poisson(link = "log"))
cat("Poisson AIC:", model_poisson_q2$aic, "\n")
```

### Model Comparison
```{r mae_all_models_q2}
# Build test model matrix and track which rows are kept
X_test <- model.matrix(
  ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station) +
    perfect_weather + semester_end + memorial_day,
  data = test_q2
)

# Align columns
X_test_aligned <- matrix(0, nrow = nrow(X_test), ncol = ncol(X_train))
colnames(X_test_aligned) <- colnames(X_train)
common_cols <- intersect(colnames(X_train), colnames(X_test))
X_test_aligned[, common_cols] <- X_test[, common_cols]

# Get row indices that model.matrix kept
kept_rows <- as.integer(rownames(X_test))

# Create full prediction vector with NA for dropped rows
pred_pois_full <- rep(NA_real_, nrow(test_q2))
pred_pois_full[kept_rows] <- as.vector(exp(X_test_aligned %*% coef(model_poisson_q2)))

test_q2 <- test_q2 %>%
  mutate(
    pred6     = predict(model6_q2, newdata = test_q2),
    pred_pois = pred_pois_full
  )

mae_all_q2 <- data.frame(
  Model = c("1. Time + Weather", "2. + Temporal Lags", "3. + Demographics",
            "4. + Station FE (Best)", "5. + Rush Hour Interaction",
            "6. Model 4 + New Features (OLS)", "7. Model 4 + New Features (Poisson)"),
  MAE = c(
    mean(abs(test_q2$Trip_Count - test_q2$pred1),     na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred2),     na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred3),     na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred4),     na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred5),     na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred6),     na.rm = TRUE),
    mean(abs(test_q2$Trip_Count - test_q2$pred_pois), na.rm = TRUE)
  ),
  RMSE = c(
    sqrt(mean((test_q2$Trip_Count - test_q2$pred1)^2,     na.rm = TRUE)),
    sqrt(mean((test_q2$Trip_Count - test_q2$pred2)^2,     na.rm = TRUE)),
    sqrt(mean((test_q2$Trip_Count - test_q2$pred3)^2,     na.rm = TRUE)),
    sqrt(mean((test_q2$Trip_Count - test_q2$pred4)^2,     na.rm = TRUE)),
    sqrt(mean((test_q2$Trip_Count - test_q2$pred5)^2,     na.rm = TRUE)),
    sqrt(mean((test_q2$Trip_Count - test_q2$pred6)^2,     na.rm = TRUE)),
    sqrt(mean((test_q2$Trip_Count - test_q2$pred_pois)^2, na.rm = TRUE))
  )
)

kable(mae_all_q2, digits = 3,
      caption = "Full Model Comparison including New Features - Q2 2025",
      col.names = c("Model", "MAE (trips)", "RMSE (trips)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```
**Interpretation**:
Adding the three new Q2-specific features in Model 6 provides a marginal improvement over Model 4, reducing MAE from 0.687 to 0.689, which is essentially no change in OLS. However, the Poisson model in Model 7 achieves the best MAE overall at 0.658, a meaningful improvement over Model 4's 0.687, confirming that modeling trip counts as a count variable better captures the distributional properties of the data than OLS. The slightly higher RMSE in Model 7 (1.052 vs 1.040) suggests the Poisson model reduces average errors but occasionally produces larger outlier predictions. Overall, the new features alone do not dramatically improve OLS performance, but the Poisson specification demonstrates that treating demand as count data is a more theoretically appropriate modeling choice for bike share prediction.


#### Critical Reflection

### Operational Implications

The final Poisson model achieves an MAE of 0.658 trips per station-hour, meaning predictions 
are off by less than one trip on average. For low-demand overnight and midday periods this is 
operationally acceptable, but during AM and PM rush hours where demand regularly exceeds 5-10 
trips per hour, an error of 0.658 can still translate into meaningful rebalancing failures. 
The most problematic scenario is systematic underprediction at high-demand Center City stations 
during peak hours, where even a small shortfall in anticipated demand could leave stations 
empty before the morning rush ends. We would recommend deploying this system under the 
following conditions: as a decision-support tool rather than a fully automated rebalancing 
system, with human operators retaining override authority; with mandatory real-time monitoring 
of stations where predicted and observed demand diverge consistently; and with a buffer 
inventory policy at historically high-error stations to absorb prediction uncertainty.

### Equity Considerations

The error analysis reveals a troubling spatial pattern: prediction errors are highest in 
wealthier, majority-white, Center City neighborhoods and lowest in lower-income, 
transit-dependent communities. While this might appear to favor equity, it reflects a 
different concern. Lower-income neighborhoods have lower average demand, so lower absolute 
errors may mask proportionally larger relative errors that still result in poor service 
reliability for communities that depend on bike share as a genuine transportation option 
rather than a recreational amenity. Deploying a system that systematically overprioritizes 
rebalancing resources toward already well-served Center City stations could deepen existing 
disparities in bike access across Philadelphia. We recommend implementing equity-weighted 
rebalancing targets that account for transit dependency and income levels, regularly auditing 
prediction performance disaggregated by neighborhood demographics, and involving community 
stakeholders in defining acceptable service levels across different parts of the city.

### Model Limitations

The model assumes that historical temporal patterns repeat consistently, an assumption that 
breaks down during special events, extreme weather, or structural shifts in travel behavior. 
The Q2 test period in June showed higher errors than the training period, suggesting the model 
does not fully generalize across the seasonal arc of spring into summer. Key missing features 
include real-time station capacity data, bike lane network changes, points of interest density, 
and granular event calendars for venues like the Mann Center, Citizens Bank Park, and 
university campuses. The panel structure also assumes station-level independence, ignoring 
spatial autocorrelation between nearby stations that could be leveraged through spatial lag 
models. With more time, we would explore gradient boosting models such as XGBoost that handle 
nonlinear interactions between weather and time more flexibly, and incorporate a spatial 
weights matrix to capture the spillover effects of demand between neighboring stations.

------------------------------------------------------------------------

                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                             